---
title: "Diagnostics: phase transitions and PDB asymptotics"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Diagnostics: phase transitions and PDB asymptotics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
set.seed(1)
```

This vignette shows the two diagnostic tools shipped with `picreg`. Both
are designed as empirical checks of the method's theoretical claims.

```{r load}
library(picreg)
```

## 1. Phase-transition analysis

`phase_transition()` answers the question: **at what sparsity level $s$
does support recovery break down?** For fixed problem geometry $(n, p)$
and signal magnitude, it varies $s$ from $0$ to `s_max` and estimates,
by Monte Carlo, the probability of exact support recovery.

Three metrics are stored simultaneously per cell:

- `exact_recovery` — $P(\hat S = S)$, the headline number.
- `tpr` — true positive rate, $|\hat S \cap S| / |S|$.
- `fdr` — false discovery rate, $|\hat S \setminus S| / \max(|\hat S|, 1)$.

### A single configuration, multiple penalties

```{r pt-single, fig.cap = "Phase transition under three penalties."}
pt <- phase_transition(
  n        = 200,
  p        = 500,
  type     = "gaussian",
  s_max    = 12,
  m        = 30,
  penalty  = c("lasso", "scad", "mcp"),
  verbose  = FALSE
)
plot(pt)
```

Reading the plot: at low $s$ all three penalties recover the support
perfectly. As $s$ grows the curves drop — SCAD and MCP typically
maintain recovery further than L1 thanks to their unbiased-large
penalisation.

```{r pt-other-metric, fig.cap = "Same data, true positive rate."}
plot(pt, metric = "tpr")
```

```{r pt-fdr, fig.cap = "Same data, false discovery rate."}
plot(pt, metric = "fdr")
```

The numerical table is in `pt$results`:

```{r pt-results}
head(pt$results)
```

### Comparing two geometries

```{r pt-geometries, fig.cap = "Effect of n on the transition."}
pt2 <- phase_transition(
  n       = c(150, 400),
  p       = c(500, 500),
  type    = "gaussian",
  s_max   = 15,
  m       = 30,
  penalty = "scad",
  verbose = FALSE
)
plot(pt2)
```

Doubling $n$ shifts the transition rightwards, as expected from
classical high-dimensional theory.

## 2. PDB asymptotic behaviour

The PDB selector has three flavours:

- `"mc_exact"` — Monte Carlo under the family's null. Family-specific,
  most accurate, slowest.
- `"mc_gaussian"` — Monte Carlo under a Gaussian approximation (CLT).
  Family-agnostic, fast.
- `"analytical"` — Bonferroni closed-form bound. Instant, conservative.

`pdb_asymptotic()` runs all three on a Gaussian design across a
grid of sample sizes, returning the simulated null distributions and
the three resulting $\hat\lambda$ values. The intent is to
**empirically validate that `"mc_gaussian"` converges to
`"mc_exact"` as $n$ grows** — which justifies using the much faster
Gaussian path in the large-$n$ regime.

### Gaussian family — sanity check

For the Gaussian family `"mc_exact"` reduces analytically to the
Gaussian null, so the two empirical distributions should coincide
at every $n$:

```{r asym-gaussian, fig.cap = "Gaussian family: mc_exact and mc_gaussian coincide.", fig.height = 6}
as_g <- pdb_asymptotic(
  n_grid  = c(100, 500, 2000),
  p       = 200,
  type    = "gaussian",
  n_simu  = 2000L,
  verbose = FALSE
)
plot(as_g)
```

The histograms overlap perfectly — this is the consistency check.

### Poisson family — non-trivial convergence

For non-Gaussian families, the two distributions are initially distinct
and converge only asymptotically. Poisson is a particularly clear case
because the count-data null is far from Gaussian at small $n$:

```{r asym-poisson, fig.cap = "Poisson family: mc_exact converges to mc_gaussian as n grows.", fig.height = 6}
as_p <- pdb_asymptotic(
  n_grid  = c(50, 200, 1000),
  p       = 200,
  type    = "poisson",
  n_simu  = 2000L,
  verbose = FALSE
)
plot(as_p)
print(as_p)
```

At $n = 50$ the Poisson `"mc_exact"` distribution (solid grey fill) sits
visibly to the right of the Gaussian step (dashed line); by $n = 1000$
they coincide. The vertical navy line marks $\hat\lambda_{\rm PDB}$ —
what `pic()` would actually use — and the dotted black line marks the
analytical Bonferroni bound.

### Practical takeaway

The diagnostic suggests when it is safe to swap `lambda_method =
"mc_exact"` for `"mc_gaussian"`, trading a small calibration loss at
small $n$ for a significant speed-up at large $n$:

```{r mc-gaussian-fit}
n <- 500; p <- 1000; s <- 5
X <- matrix(rnorm(n * p), n, p)
beta <- numeric(p); beta[sample.int(p, s)] <- 3
y <- as.numeric(X %*% beta + rnorm(n))

fit_exact    <- pic(X, y, family = "gaussian", lambda_method = "mc_exact")
fit_gaussian <- pic(X, y, family = "gaussian", lambda_method = "mc_gaussian")

c(exact = fit_exact$lambda, gaussian = fit_gaussian$lambda)
setequal(fit_exact$selected, fit_gaussian$selected)
```

## Summary

| Diagnostic | Question answered | Use when |
|---|---|---|
| `phase_transition()` | "How sparse can my problem be and still be recoverable?" | Benchmarking, paper figures |
| `pdb_asymptotic()` | "Can I use the fast Gaussian PDB instead of the exact one?" | Tuning, regime check |
