---
title: "Getting started with picreg"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with picreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
set.seed(1)
```

## What `picreg` does

`picreg` fits sparse generalised linear models in high dimensions. Unlike
`glmnet`, the regularisation parameter $\lambda$ is **not** chosen by
cross-validation: it is computed analytically (or by short Monte Carlo)
from the **Pivotal Detection Boundary** (PDB) — the $(1-\alpha)$
quantile of a pivotal null-distribution statistic. One fit, no
hyperparameter tuning loop.

The package supports six families and three penalties:

| Family | Continuous / discrete | Response format |
|---|---|---|
| `"gaussian"` | continuous | `y` numeric |
| `"binomial"` | binary | `y` in $\{0, 1\}$ |
| `"poisson"` | counts | `y` non-negative integer |
| `"exponential"` | positive continuous | `y` positive |
| `"gumbel"` | continuous (heavy upper tail) | `y` numeric |
| `"cox"` | survival | `y = cbind(time, event)` |

Penalties: `"lasso"`, `"scad"` (with `scad_a`), `"mcp"` (with `mcp_gamma`).

```{r load}
library(picreg)
```

## First fit — Gaussian + SCAD

We generate a sparse design with $n = 200$, $p = 500$, $s = 5$ true
non-zero coefficients of magnitude $3$:

```{r gaussian-fit}
n <- 200; p <- 500; s <- 5
X <- matrix(rnorm(n * p), n, p)
true_features <- sort(sample.int(p, s))
beta <- numeric(p); beta[true_features] <- 3
y <- as.numeric(X %*% beta + rnorm(n))

fit <- pic(X, y, family = "gaussian", penalty = "scad")
fit
```

The selected support and the PDB-chosen $\lambda$:

```{r gaussian-inspect}
fit$selected
fit$lambda

# Recovery check
setequal(fit$selected, true_features)
```

### Visualising the coefficient pattern

```{r gaussian-plot, fig.cap = "Non-zero coefficients of the SCAD fit."}
plot(fit)
```

### Visualising the PDB null distribution

The histogram of the simulated null gradient-norm statistics, with a
vertical line at the selected $\hat\lambda$:

```{r gaussian-lambda-plot, fig.cap = "PDB null distribution and selected lambda."}
plot(fit$lambda_pdb)
```

A descriptive summary of the null distribution is available via
`pdb_summary()`:

```{r gaussian-pdb-summary}
pdb_summary(fit)
```

## Logistic regression

```{r binomial-fit}
eta  <- X %*% beta
p_y  <- 1 / (1 + exp(-eta))
y_b  <- rbinom(n, 1, p_y)

fit_bin <- pic(X, y_b, family = "binomial", penalty = "lasso")
fit_bin
setequal(fit_bin$selected, true_features)
```

Prediction returns the response-scale probability by default; pass
`type = "link"` for the linear predictor or `type = "class"` for
hard $\{0, 1\}$ labels:

```{r binomial-predict}
head(predict(fit_bin, X, type = "response"))
head(predict(fit_bin, X, type = "class"))
```

## Switching penalty without rerunning PDB

The 4-step path always starts with three warm-up steps under L1 and
switches to the requested penalty only at the final step. Setting
`penalty = "lasso"`, `"scad"`, or `"mcp"` changes only the last step
— the lambda selector is identical:

```{r penalty-comparison}
fits <- lapply(c("lasso", "scad", "mcp"),
               function(pen) pic(X, y, family = "gaussian", penalty = pen))
sapply(fits, function(f) length(f$selected))
sapply(fits, function(f) f$lambda)   # identical (same PDB call)
```

## Where to go next

- [Cox regression with picreg] — survival analysis, baseline functions,
  c-index.
- [Diagnostics: phase transitions and PDB asymptotics] — empirical
  validation of the method on synthetic data.
- `?pic` and `?lambda_pdb` for the full argument reference.
