---
title: "Cox regression with picreg"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Cox regression with picreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
set.seed(1)
```

## Setting

The Cox family fits a partial-likelihood model (Breslow tie handling).
Compared to other families it has two specificities:

1. **No intercept** is fitted (`intercept` is forced to `FALSE`).
2. The response `y` must be a two-column matrix `cbind(time, event)`,
   where `event` is `1` for an observed event and `0` for censoring.

The PDB selector for Cox uses a custom permutation-based pivotal
estimator (implemented in `src/cox_null_grad.cpp`) for the
`"mc_exact"` method. The `"mc_gaussian"` and `"analytical"` methods
fall back to the family-agnostic paths.

```{r load}
library(picreg)
```

## Simulated survival data

We simulate event times under an exponential proportional-hazards
model with rate $\exp(X\beta)$ and uniform censoring:

```{r simulate}
n <- 300; p <- 200; s <- 5
X <- matrix(rnorm(n * p), n, p)
true_features <- sort(sample.int(p, s))
beta <- numeric(p); beta[true_features] <- 0.7

eta     <- as.numeric(X %*% beta)
t_event <- rexp(n, rate = exp(eta))
t_cens  <- rexp(n, rate = 0.5)
time    <- pmin(t_event, t_cens)
event   <- as.integer(t_event <= t_cens)
y_surv  <- cbind(time = time, event = event)

mean(event)   # observed event rate
```

## Fitting the model

```{r cox-fit}
fit <- pic(X, y_surv, family = "cox", penalty = "lasso")
fit
fit$selected
setequal(fit$selected, true_features)
```

The fit stores three Cox-specific quantities estimated after the
penalised fit:

```{r cox-attributes}
head(fit$baseline_cumulative_hazard)
head(fit$baseline_survival)
fit$censoring_rate    # percentage of censored observations
```

## Baseline functions

```{r cox-baseline-plot, fig.height = 5, fig.cap = "Breslow baseline cumulative hazard and baseline survival."}
plot_baseline(fit)
```

## Performance metric: Harrell's C-index

The standalone helper `concordance_index()` computes Harrell's
C-index given a fitted linear predictor:

```{r cox-cindex}
linpred <- as.numeric(X %*% fit$beta)
concordance_index(time, event, linpred)
```

A C-index of $0.5$ indicates random prediction; $1$ is perfect
ordering. Values around $0.7$–$0.8$ are common on real-world clinical
data.

## Survival curves for new patients

`predict_survival_function()` returns $S(t \mid x)$ for one or more
new covariate vectors:

```{r cox-predict}
new_X <- X[1:3, , drop = FALSE]
surv  <- predict_survival_function(fit, new_X)
str(surv)
```

The returned list has `time` (event times where the curve drops) and
`survival` (a matrix with one column per new individual).

## Cross-checking with the partial log-likelihood

```{r cox-pll}
cox_partial_log_likelihood(time, event, linpred)
```

Higher values indicate better fit; this is the same quantity
maximised internally during the FISTA optimisation.

## When to use `"mc_gaussian"` instead of `"mc_exact"`

The default `"mc_exact"` runs a permutation Monte Carlo that scales
with `n_simu × n`. For large $n$, the much faster `"mc_gaussian"`
approximation is well-calibrated — see the *Diagnostics* vignette
for the empirical convergence picture.
